Towards topological type theory for decrypting transfinite methods in classical mathematics

### 2025-08-25 slides-ai2025.tex
Rosetta stone and test cases

\begin{frame}{A glimpse of proof mining}
  \textbf{Theorem.} For every natural number~$n$, there is a prime number
  larger than~$n$.~\fixedhref{https://unimath.github.io/agda-unimath/elementary-number-theory.infinitude-of-primes.html}{\formalized}

@@

  \begin{mining}
    \textbf{Scholium.} $p_n \leq \lceil e^{n+1-\gamma} \rceil$.
  \end{mining}
\end{frame}

% 1995: conjecture
% 2003: solution
% 2018: rate of convergence
\begin{frame}{Proof mining in convex analysis}
  \textbf{Theorem (zero displacement conjecture).}
  Let~$H$ be a Hilbert space. \newline
  Let~\mbox{$C_1, \ldots, C_N \subseteq H$} be nonempty
  closed convex subsets with orthogonal projections~$P_{C_i}$. Let~$T = P_{C_N}
  \circ \ldots \circ P_{C_1}$. Then for every~$x \in H$,
  \[ \| T^{n+1}x - T^nx \| \xrightarrow{n \to \infty} 0. \]

@@

\begin{frame}{Proof mining in approximation theory}
  Let~$n \in \NN$. Let~$P_n$ be the space of polynomials of degree at most~$n$. \\
  Let~$f \in C[0,1]$ be a continuous function.

@@

  \begin{mining}[0.85\textwidth]
    \textbf{Scholium
    \fixedhref{https://www2.mathematik.tu-darmstadt.de/~kohlenbach/Keele.pdf}{[Kohlenbach
    1990]}.} $\forall \varepsilon > 0\_ \forall p_1,p_2 \in P_n\_$
    \[
      \bigwedge_{i=1}^2 \bigl(\|f - p_i\|_1 - \operatorname{dist}_1(f, P_n) <
      \phi(\omega,n,\varepsilon)\bigr) \Longrightarrow \|p_1 - p_2\|_1 \leq
      \varepsilon, \]
    where~$\phi(\omega,n,\varepsilon)$ is given by a certain explicit formula.
  \end{mining}
\end{frame}

\begin{frame}{Backed by logical metatheorems?}
  \hil{Metatheorems} have been and are being developed which \ldots
  \begin{enumerate}
    \item guarantee the \hil{extractability} of suitable numerical information in principle
    (bounds on norms, convergence rates, moduli of uniqueness, rates of
    asymptotic regularity, \ldots),
    \item describe an \hil{algorithm} for carrying out the extraction and
    \item support \hil{modular treatments} of auxiliary lemmas,
  \end{enumerate}
  \bad{\textbf{provided}} the input proof is supplied in a
  certain \hil{formal system}.

  NB: The \hil{quality} of the extracted data depends on the
  \hil{sophistication} of the \hil{logical~principles} used in the proof
  (Heine--Borel, Bolzano--Weierstraß, \ldots).

  \bad{\textbf{In practice (2025):}} Algorithms \bad{\textbf{not}} used, only followed
  as \bad{\textbf{rough guidelines}}, combined with hand-rolled optimizations.
\end{frame}

>>>>

\NewEnviron{simpleblock}{
  \colorbox{mypurple!10}{\begin{minipage}{\textwidth}
    \BODY
  \end{minipage}}
}

\renewcommand{\insertframeextra}{a}
{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.3cm}\includegraphics[width=\paperwidth]{fr1-lighter}\end{minipage}}
\begin{frame}{A first glimpse of proof mining}
  \textbf{Theorem.} For every natural number~$n$, there is a prime number
  larger than~$n$.~\fixedhref{https://unimath.github.io/agda-unimath/elementary-number-theory.infinitude-of-primes.html}{\formalized}

@@

  \begin{mining}
    \textbf{Scholium.} $p_n \leq \lceil e^{n+1-\gamma} \rceil$.
  \end{mining}
\end{frame}}

% 1995: conjecture
% 2003: solution
% 2018: rate of convergence
\addtocounter{framenumber}{-1}
\renewcommand{\insertframeextra}{b}
\begin{frame}{Proof mining in convex analysis}
  \textbf{Theorem (zero displacement conjecture).}
  Let~$H$ be a Hilbert space. \newline
  Let~\mbox{$C_1, \ldots, C_N \subseteq H$} be nonempty
  closed convex subsets with orthogonal projections~$P_{C_i}$. Let~$T = P_{C_N}
  \circ \ldots \circ P_{C_1}$. Then for every~$x \in H$,
  \[ \| T^{n+1}x - T^nx \| \xrightarrow{n \to \infty} 0. \]

@@

\addtocounter{framenumber}{-1}
\renewcommand{\insertframeextra}{c}
\begin{frame}{Proof mining in approximation theory}
  Let~$n \in \NN$. Let~$P_n$ be the space of polynomials of degree at most~$n$. \\
  Let~$f \in C[0,1]$ be a continuous function.

@@

  \begin{mining}[0.85\textwidth]
    \textbf{Scholium
    \fixedhref{https://www2.mathematik.tu-darmstadt.de/~kohlenbach/Keele.pdf}{[Kohlenbach
    1990]}.} $\forall \varepsilon > 0\_ \forall p_1,p_2 \in P_n\_$
    \[
      \bigwedge_{i=1}^2 \bigl(\|f - p_i\|_1 - \operatorname{dist}_1(f, P_n) <
      \phi(\omega,n,\varepsilon)\bigr) \Longrightarrow \|p_1 - p_2\|_1 \leq
      \varepsilon, \]
    where~$\phi(\omega,n,\varepsilon)$ is given by a certain explicit formula.
  \end{mining}
\end{frame}
\renewcommand{\insertframeextra}{}

\begin{frame}{Backed by logical metatheorems?}
  \hil{Metatheorems} have been and are being developed which \ldots
  \begin{enumerate}
    \item guarantee the \hil{extractability} of suitable numerical information in principle
    (bounds on norms, convergence rates, moduli of uniqueness, rates of
    asymptotic regularity, \ldots),
    \item describe an \hil{algorithm} for carrying out the extraction and
    \item support \hil{modular treatments} of auxiliary lemmas,
  \end{enumerate}
  \bad{\textbf{provided}} the input proof is supplied in a
  certain \hil{formal system}.
  \pause

  NB: The \hil{quality} of the extracted data depends on the
  \hil{sophistication} of the \hil{logical~principles} used in the proof
  (Heine--Borel, Bolzano--Weierstraß, \ldots).
  \pause

  \bad{\textbf{In practice (2025):}} Algorithms \bad{\textbf{not}} used, only followed
  as \bad{\textbf{rough guidelines}}, combined with hand-rolled optimizations.
  \hil{Let us explore tool support!}
\end{frame}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\centering\vspace*{-0em}\includegraphics[width=\paperwidth]{curry-howard-faded}\end{minipage}}
\newcommand{\connection}[1]{$\cdots\includegraphics[height=1em]{#1}\cdots$}
\begin{frame}{Curry--Howard's Rosetta stone}
  \bigskip
  \centering
  \renewcommand{\arraystretch}{1.3}
  \begin{tabular}{r@{}c@{}l}
    \hil{mathematics} & & \hil{programming} \\
    proving a claim & \connection{emoji-thinking} & implementing a function \\
    stating a claim & \connection{emoji-megaphone} & writing down a type signature \\
    using a lemma & \connection{emoji-link} & calling a function \\
    {\ \ \ \ \,}coming up with an auxiliary lemma & \connection{emoji-lightbulb} & coming up with an auxiliary function \\
    committing a logical fallacy & \connection{emoji-cross} & committing a type error \\
    \pause
    --- & \connection{emoji-explosion} & encountering a runtime bug
  \end{tabular}
  \pause

  \begin{hilblock}
    \centering
    To extract numerical data from a proof, \\
    \hil{run the proof}.
  \end{hilblock}
\end{frame}}

\begin{frame}{Three case studies}
  \bigskip
  \begin{columns}[c]
    \begin{column}{0.15\textwidth}
      \scalebox{0.23}{\sieve{7}{7}}
    \end{column}
    \begin{column}{0.85\textwidth}
      \begin{simpleblock}
        \justifying
        \textbf{Theorem.}
        For every natural number~$n$, there is a prime larger than~$n$.
      \end{simpleblock}
      \\[0.3em]

      \emph{Proof.} Every prime factor of~$n! + 1$ will do. \qed
    \end{column}
  \end{columns}
  \bigskip

  \begin{columns}[c]
    \begin{column}{0.15\textwidth}
      \includegraphics[width=\textwidth,valign=t]{monad}
    \end{column}
    \begin{column}{0.85\textwidth}
      \begin{simpleblock}
        \textbf{Theorem.}
        Every infinite sequence~$f : \NN \to \NN$ is \emph{good}
        in that there are numbers $i < j$ such that~$f(i) \leq f(j)$.
      \end{simpleblock}
      \\[0.3em]

      \emph{Proof.} There is a \bad{minimal value}~$f(i)$. Set~$j \defeq i+1$. \qed
    \end{column}
  \end{columns}
  \bigskip

  \begin{columns}[c]
    \begin{column}{0.15\textwidth}
      \includegraphics[width=\textwidth,valign=t]{calabi-yau}
    \end{column}
    \begin{column}{0.85\textwidth}
      \begin{simpleblock}
        \justifying
        \textbf{Theorem.}
        Let~$M$ be a surjective matrix with more rows than columns over a
        commutative ring~$A$. Then~$1 = 0$ in~$A$.
      \end{simpleblock}
      \\[0.3em]

      \justifying\emph{Proof.}
      \bad{Assume not.} Then there is~a \bad{maximal ideal} $\mmm$.
      The matrix is still surjective over~$A/\mmm$. Since~$A/\mmm$ is a field, this
      is a contradiction to basic linear algebra. \qed
    \end{column}
  \end{columns}
\end{frame}
